Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/appendix.tex
clean up trie exposition

Our last example generates a trie intermediate data structure.

The function {\tt exists} below determines whether a query string (s) occurs in a given set of strings (from the alphabet $\{A, B,C\}$).
It solves this problem by partitioning the provided set of strings into three subsets,
one which contains all words that start with $A$,
one which contains all words that start with $B$, and one for $C$.
When partitioning, we also throw away the first letter, since we know it's uniform for the whole subset.
Also while doing this partitioning, we also keep track of whether the empty word occurred at all.
We then look at the query string.  If it's empty, we return whether the empty word was in the set.
Otherwise, we look at the first letter, pick the appropriate subset, 
and recursively search for the rest of our query string in that.

We stage this algorithm by saying that the word set comes at stage \bbone\
and the query string comes at stage \bbtwo.
\begin{lstlisting} 
atsignmono { 
datatype letter = A | B | C 
datatype string = EmptyS | ConsS of letter * string
datatype list   = EmptyL | ConsL of string * list

@@

fun exists (mono{EmptyL},next{s}) = next{false} 
  | exists (mono{ConsL ht},next{s}) = 
	let atsignmono{
	  val (anyEmpty,a,b,c) = partition (ConsL ht)
	} in next { 
	case s of 
	  EmptyS2 => prev{hold mono{anyEmpty}} 
	| ConsS2 (A2,z) => prev{exists(mono{a},next{z})} 
	| ConsS2 (B2,z) => prev{exists(mono{b},next{z})}
	| ConsS2 (C2,z) => prev{exists(mono{c},next{z})}  
	} 
\end{lstlisting}
Running the splitting algorithm on this code produces:
\begin{lstlisting} 
datatype trie = Leaf 
			  | Branch of bool * trie * trie * trie

@@

As promised, this split version creates a trie in the first stage, and consumes it in the second stage.

>>>>

The function {\tt exists} determines whether a query string \texttt{s} occurs in
a given list of strings (over the alphabet $\{A,B,C\}$). It starts by
partitioning the provided set of strings into three subsets, corresponding to
the set of strings starting with $A$ (resp., $B$ and $C$), and stripping that
first letter off of each of those strings. While processing the input list, it
also keeps track of whether the empty string has been encountered.
%
We then look at the query string \texttt{s}. If it's empty, we return whether
the empty string was encountered. Otherwise, we choose the partition which
agrees with the first letter of \texttt{s}, and recursively searches that list
of strings for the remainder of \texttt{s}.

Our staging annotations indicate that the list of strings are first-stage, while
the query string is second-stage.
\begin{lstlisting} 
atsigngr { 
datatype letter = A | B | C 
datatype string = EmptyS | ConsS of letter * string
datatype list   = EmptyL | ConsL of string * list

@@

fun exists (gr{EmptyL},next{s}) = next{false} 
  | exists (gr{ConsL ht},next{s}) = 
	let atsigngr{
	  val (anyEmpty,a,b,c) = partition (ConsL ht)
	} in next { 
	case s of 
	  EmptyS2 => prev{hold gr{anyEmpty}} 
	| ConsS2 (A2,z) => prev{exists(gr{a},next{z})} 
	| ConsS2 (B2,z) => prev{exists(gr{b},next{z})}
	| ConsS2 (C2,z) => prev{exists(gr{c},next{z})}  
	} 
\end{lstlisting}
Running the splitting algorithm on this code produces:
\begin{lstlisting} 
datatype trie = Leaf 
			  | Branch of bool * trie * trie * trie

@@

\texttt{exists1} creates a trie (or prefix tree) as the boundary data structure,
which is traversed by \texttt{exists2}.
